Type-Safe Modular Parsing

### 2016-12-22 paper/sections/ObjectAlgebras.tex
Code format

Object Algebras, first introduced by ... et al as a solution to the famous \textit{Expression Problem}, provide extensibility on
both data variants and operations for structures like abstract syntax trees. We integrate Object Algebras directly to enhance the
extensibility of open parsers.

@@

In Object Algebras, ASTs as recursive data structures are defined using traits, where each constructor corresponds
to an abstract method inside. The example from Section~\ref{subsec:parsingwithopen} is again used here for illustration.
At first the language only supports variables and applications:
\begin{lstlisting}
trait ExprAlg[E] {
    def varE(x : String) : E
    def appE(e1 : E, e2 : E) : E
}
\end{lstlisting}
The language has two constructors: literals and additions. The trait \lstinline{ExprAlg} is called an \textit{Object Algebra interface},
whereas as a factory, it cannot produces objects for the expressions directly like traditional approaches, but abstract the results in its
type parameter \lstinline{E} instead. To realize an operation on expressions, we simply instantiate the type parameter by a concrete type and
provides implementations for all cases. Below is an example of collecting all free variables in an expression:
\begin{lstlisting}
trait FreeVars extends ExprAlg[List[String]] {
    def varE(x : String) = List(x)
    def appE(e1 : List[String], e2 : List[String]) = e1 ++ e2
}
\end{lstlisting}
Here \lstinline{FreeVars} is called an \textit{Object Algebra}. It traverses an expression bottom-up, and returns a list of strings as the result.
Furthermore, one can define a new trait to implement pretty-printing on expressions:
\begin{lstlisting}
trait ExprPrint extends ExprAlg[String] {
    def varE(x : String) = x
    def appE(e1 : String, e2 : String) = "(" + e1 + " " + e2 + ")"
}
\end{lstlisting}
On the other hand, the data variants can be extended by inheriting \lstinline{ExprAlg} and adding new cases only. Suppose we want to
have literals in expressions, a new Object Algebra interface \lstinline{LitAlg} is defined as follows:
\begin{lstlisting}
trait LitAlg[E] extends ExprAlg[E] {
    def litE(n : Int) : E
}
\end{lstlisting}

@@

\begin{lstlisting}
trait LitPrint extends ExprPrint {
    def litE(n : Int) = n.toString
}
\end{lstlisting}

@@

\begin{lstlisting}
def build[E](alg : LitAlg[E]) : E =
    alg.appE(alg.varE("x"), alg.litE(3))
\end{lstlisting}

@@

\begin{lstlisting}
build(new LitPrint{})
\end{lstlisting}
results in \lstinline{"(x 3)"}, as the result of pretty-printing. Now we observe that an ``object'' of expression actually
has the function type \lstinline{LitAlg[E] =>} \lstinline{E} for generic \lstinline{E}. At this point, one may argue
that it is illusory compared to traditional objects, and since we are struggling against the parsing problem, if a parser
produces structures as functions, it could be hardly be further processed like objects, especially, in the design of a compiler
there will be a lot of desugarings (or transformations). However, we argue that this is not an issue, because functions can be used
in a very clever way. For example, if one wants to transform an \lstinline{ExprAlg} expression to a \lstinline{LitAlg} expression by
replacing some variables with values based on a variable environment, a transformation algebra can be implemented as follows:
\begin{lstlisting}
trait Refactor[E] extends ExprAlg[E] {
    def alg : SubAlg[E]
    def env : Map[String, Int]
    def varE(x : String) = if (env.contains(x)) alg.litE(env(x)) else alg.varE(x)
}
\end{lstlisting}\haoyuan{Is this example OK?}
When \lstinline{env = Map("y"} \lstinline{-> 3)}, the \lstinline{ExprAlg} expression \lstinline{"x y"} is changed to \lstinline{LitAlg} expression \lstinline{"x 3"}. The abstract method
\lstinline{alg} in \lstinline{Refactor} can be any algebra of \lstinline{SubAlg}, simply for delegation. By passing a concrete implementation
to \lstinline{alg}, \lstinline{Refactor} will be able to apply transformation to an \lstinline{ExprAlg} expression before \lstinline{alg} returns the result.

>>>>

Object Algebras~\cite{Oliveira2012} is a solution to the famous \textit{Expression Problem}, provide extensibility on
both data variants and operations. It is suitable for structures like abstract syntax trees. We integrate Object Algebras directly to enhance the extensibility of \name.

@@

In Object Algebras, ASTs as recursive data structures are defined using traits, where each constructor corresponds
to an abstract method inside. The example from Section~\ref{subsec:parsingwithopen} is again used here for illustration.
At first the language only supports variables and applications:

\lstinputlisting[linerange=4-8]{../Scala/Parser/src/PaperCode/ObjectAlg/OA1.scala}% APPLY:linerange=OBJECT_ALG_EXPR

The language has two constructors: literals and additions. The trait \lstinline{ExprAlg} is called an \textit{Object Algebra interface},
whereas as a factory, it cannot produces objects for the expressions directly like traditional approaches, but abstract the results in its
type parameter \lstinline{E} instead. To realize an operation on expressions, we simply instantiate the type parameter by a concrete type and
provides implementations for all cases. Below is an example of collecting all free variables in an expression:

\lstinputlisting[linerange=12-15]{../Scala/Parser/src/PaperCode/ObjectAlg/OA1.scala}% APPLY:linerange=OBJECT_ALG_FREE_VARS

Here \lstinline{FreeVars} is called an \textit{Object Algebra}. It traverses an expression bottom-up, and returns a list of strings as the result.
Furthermore, one can define a new trait to implement pretty-printing on expressions:

\lstinputlisting[linerange=19-22]{../Scala/Parser/src/PaperCode/ObjectAlg/OA1.scala}% APPLY:linerange=OBJECT_ALG_EXPR_PRINT

On the other hand, the data variants can be extended by inheriting \lstinline{ExprAlg} and adding new cases only. Suppose we want to
have literals in expressions, a new Object Algebra interface \lstinline{LitAlg} is defined as follows:

\lstinputlisting[linerange=26-28]{../Scala/Parser/src/PaperCode/ObjectAlg/OA1.scala}% APPLY:linerange=OBJECT_ALG_LIT

@@

\lstinputlisting[linerange=32-34]{../Scala/Parser/src/PaperCode/ObjectAlg/OA1.scala}% APPLY:linerange=OBJECT_ALG_LIT_PRINT

@@

\huang{todo: better name for build}
\lstinputlisting[linerange=48-48]{../Scala/Parser/src/PaperCode/ObjectAlg/OA1.scala}% APPLY:linerange=OBJECT_ALG_BUILD

@@

\begin{lstlisting}
build(new LitPrint{})
\end{lstlisting}

results in \lstinline{"(x 3)"}, as the result of pretty-printing. Now we observe that an ``object'' of expression actually
has the function type \lstinline{LitAlg[E] =>} \lstinline{E} for generic \lstinline{E}. At this point, one may argue
that it is illusory compared to traditional objects, and since we are struggling against the parsing problem, if a parser
produces structures as functions, it could be hardly be further processed like objects, especially, in the design of a compiler
there will be a lot of desugarings (or transformations). However, we argue that this is not an issue, because functions can be used
in a very clever way. For example, if one wants to transform an \lstinline{ExprAlg} expression to a \lstinline{LitAlg} expression by
replacing some variables with values based on a variable environment, a transformation algebra can be implemented as follows:

\lstinputlisting[linerange=38-42]{../Scala/Parser/src/PaperCode/ObjectAlg/OA1.scala}% APPLY:linerange=OBJECT_ALG_REFACTOR

When \lstinline{env = Map("y"} \lstinline{-> 3)}, the \lstinline{ExprAlg} expression \lstinline{"x y"} is changed to \lstinline{LitAlg} expression \lstinline{"x 3"}. The abstract method
\lstinline{alg} in \lstinline{Refactor} can be any algebra of \lstinline{SubAlg}, simply for delegation. By passing a concrete implementation
to \lstinline{alg}, \lstinline{Refactor} will be able to apply transformation to an \lstinline{ExprAlg} expression before \lstinline{alg} returns the result.
